---
title: "RNA-Seq_SRA"
output: html_document
date: "2025-05-26"
---


```{r}

library(DESeq2)

library(apeglm)

library(dplyr)

library(tidyverse)

library(ggplot2)

library(ggrepel)

```


```{r}

setwd("~/Documents/GitHub/KRAS-RNA Seq/RNA-Seq-mini-project-KRAS/")

df_kras <- read.delim("~/Documents/GitHub/KRAS-RNA Seq/RNA-Seq-mini-project-KRAS/Sample Set/GSE295712_gene-featureCount-else-int_name.txt")


head(df_kras[, 1:5])

counts_agg <- aggregate(. ~ GeneName, data = df_kras, FUN = sum)

rownames(counts_agg) <- counts_agg$GeneName

counts <- counts_agg[, -1]

#Specify the contrast groups

condition <- factor(c("Lung", "Lung", "Lymph", "Lymph", "Lymph"))

coldata <- data.frame(condition = condition)

rownames(coldata) <- colnames(counts)

#Run DESEq2

dds <- DESeqDataSetFromMatrix(countData = counts, colData = coldata, design = ~ condition)

dds <- DESeq(dds)

#Normalized the counts data

normalized_counts <- counts(dds, normalized=TRUE)

#Specify the levels of contrast

res_kras <- results(dds, contrast = c("condition","Lymph","Lung"))

res_kras

#Remove all NA fields

res_kras <- na.omit(res_kras)

res_kras

res_kras$gene <- rownames(res_kras)

res_kras <- res_kras[, c("gene", setdiff(names(res_kras), "gene"))]

res_kras

res_kras_df <- as.data.frame(res_kras)

write.csv(res_kras_df, file = "~/Documents/GitHub/kras_df.csv")

```

#Running Pre-ranked GSEA

```{r}
res_kras_df <- res_kras_df %>% 
  mutate(rnk = sign(log2FoldChange) * (-log10(pvalue)))

subset_res_kras <- res_kras_df %>%
  select(rnk)

subset_res_kras <- arrange(subset_res_kras, -rnk)

write.csv(subset_res_kras, file = "~/Documents/GitHub/KRAS-RNA Seq/RNA-Seq-mini-project-KRAS/files for GSEA/rnk.csv") #Save the file as .rnk after removing the header from the first row. I ran the GSEA from the desktop app.

```

#Visualizing the up and downregulated pathways from GSEA output

```{r}
neg_kras <- read_tsv("~/Documents/GitHub/KRAS-RNA Seq/RNA-Seq-mini-project-KRAS/files for GSEA/kras.GseaPreranked.1748227258719/gsea_report_for_na_neg_1748227258719.tsv")

head(neg_kras)

neg_df <- neg_kras[, c("NAME", "FDR q-val")] %>%
  mutate(log10fdr = log10(`FDR q-val`+1e-10)) #Needed to add a value to the FDR values=0


pos_kras <- read_tsv("~/Documents/GitHub/KRAS-RNA Seq/RNA-Seq-mini-project-KRAS/files for GSEA/kras.GseaPreranked.1748227258719/gsea_report_for_na_pos_1748227258719.tsv")

head(pos_kras)

pos_df <- pos_kras[, c("NAME", "FDR q-val")] %>%
  mutate(log10fdr = -log10(`FDR q-val`+1e-10))


merge_kras <- full_join(neg_df, pos_df)


merge_kras <- filter(merge_kras, log10fdr > 1.3 | log10fdr < -1.3)

merge_kras <- merge_kras %>%
  mutate(cond = case_when(
    log10fdr > -1.3 ~ "red",
    log10fdr < 1.3 ~"blue"
  ))


```

#Visualizing the pathways plot - the plot is shown in Figures folder

```{r}
plot_kras <- ggplot(merge_kras, mapping = aes(x=log10fdr, y=reorder(NAME, -log10fdr))) + 
  geom_bar(stat = "identity", color = "black", aes(fill=cond)) + 
  scale_fill_identity() +
  xlab("-log10(FDR q-value)") +
  ylab("Hallmark Pathways") +
  geom_vline(xintercept = 1.3, linetype = "dashed", color = "black") +
  geom_vline(xintercept =  -1.3, linetype = "dashed", color = "black")

plot_kras

```
#Visualizing gene level analysis from GSEA output. The plot is shown in the Figures folder

```{r}
res_kras_df$Pathways <- "NS"
res_kras_df$Pathways[res_kras_df$log2FoldChange > 1.3 & res_kras_df$pvalue < 0.05] <- "Up" #Based on the approrpiate cutoffs
res_kras_df$Pathways[res_kras_df$log2FoldChange < -1.3 & res_kras_df$pvalue < 0.05] <- "Down"

top50_kras <- head(res_kras_df[order(res_kras_df$pvalue), "gene"], 50)


res_kras_df$de50kras <- ifelse(res_kras_df$gene %in% top50_kras, res_kras_df$gene, NA)


volcano_kras <- ggplot(res_kras_df, aes(x=log2FoldChange, y=-log10(pvalue), col = Pathways, label = de50kras)) +
  geom_point() +
  geom_vline(xintercept = 1.3, linetype = "dashed", color = "black") +
  geom_vline(xintercept = -1.3, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = -1.3, linetype = "dashed", color = "blue") +
  scale_color_manual(values = c("red", "grey", "blue"),
                     labels = c("Downregulated", "Not Significant", "Upregulated")) +
  geom_text_repel(max.overlaps = Inf) 

volcano_kras
```

#Pathway specific volcano plots -- here I have shown the example of Angiogenesis pathway since the authors found VEGFA being upregulated in the lymph nodes compared to lung samples. The plot is shown in the Figures folder.


```{r}

ang <- read.csv("~/Documents/GitHub/KRAS-RNA Seq/RNA-Seq-mini-project-KRAS/Sample set/Angiogenesis.csv")


ang$Pathways <- "NS"
ang$Pathways[ang$log2FoldChange > 1.3 & ang$pvalue < 0.05] <- "Up"
ang$Pathways[ang$log2FoldChange < -1.3 & ang$pvalue < 0.05] <- "Down"

top10_ang <- head(ang[order(ang$pvalue), "gene"], 10)

ang$de10ang <- ifelse(ang$gene %in% top10_ang, ang$gene, NA)


volcano_ang <- ggplot(ang, aes(x=log2FoldChange, y=-log10(pvalue), col = Pathways, label = de10ang)) +
  geom_point() +
  geom_vline(xintercept = 1.3, linetype = "dashed", color = "black") +
  geom_vline(xintercept = -1.3, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = -1.0, linetype = "dashed", color = "blue") +
  scale_color_manual(values = c("red", "grey", "blue"),
                     labels = c("Downregulated", "Not Significant", "Upregulated")) +
  geom_text_repel(max.overlaps = Inf) 

volcano_ang
```
